CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-04-08 script/en/chapters/aboutcryptool.tex
be updates

You will obtain information about the principles of symmetrical and
asymmetrical {\bf encryption}. A large section of this script is dedicated
to the fascinating topic of {\bf prime numbers}. Using numerous examples,
the {\bf elementary number theory} and {\bf modular arithmetic} are
introduced and applied in an exemplary manner for the {\bf RSA procedure}.
By reading this you'll gain an insight into the mathematical ideas behind
{\bf modern cryptography}.

@@

The {\em authors} Bernhard Esslinger, Matthias B\"uger, Bartol Filipovic, 
Henrik Koy, Roger Oyono and J\"org-Cornelius Schneider
would like to take this opportunity to thank their colleagues 
in the company and at the universities of Frankfurt, Gie\ss en, 
Siegen, Karlsruhe and Darmstadt. They are particularly indepted Dr. Peer
Wichmann from the Karlsruhe computer science research centre (Forschungszentrum
Informatik, FZI) for his down-to-earth support.

>>>>

You will obtain information about the principles of symmetrical and
asymmetrical {\bf encryption}. A large section of this script is dedicated
to the fascinating topic of {\bf prime numbers}. Using numerous examples,
the {\bf elementary number theory} and {\bf modular arithmetic} are
introduced and applied in an exemplary manner for the {\bf RSA procedure}.
By reading the following chapter you'll gain an insight into the mathematical
ideas behind {\bf modern cryptography}.

@@

The {\em authors} Bernhard Esslinger, Matthias B\"uger, Bartol Filipovic, 
Henrik Koy, Roger Oyono and J\"org-Cornelius Schneider
would like to take this opportunity to thank their colleagues 
in the company and at the universities of Frankfurt, Gie\ss en, 
Siegen, Karlsruhe and Darmstadt. They are particularly indepted to Dr.~Peer
Wichmann from the Karlsruhe computer science research centre (Forschungszentrum
Informatik, FZI) for his down-to-earth support.

### 2003-04-08 script/en/chapters/cryptomethods.tex
cvs id
be updates

% ..............................................................................
%                V E R S C H L U E S S E L U N G S V E R F A H R E N
%
% Schreibregelung: Capitel header with all nouns/verbs-words capitalized,
%                  All other headers in the normal lower/upper case manner
%                  like sentences without dot at the end.
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

In their 2001-paper Ferguson, Schroeppel and Whiting \cite{Ferguson2001}
presented a new method of symmetric codes cryptanalysis: They described AES with
a closed formula (in the form of a continued fraction) which was possible
because of the "relatively" clear structure of AES. This formula consists of
around 1000 trillion terms of a sum - so it does not help concrete practical
cryptanalysis. Nevertheless -curiosity in the academic community was awakened.
It was already known, that the 128-bit AES could be described as an
over-determined system of about 8000 quadratic equations (over an algebraic
number field) with about 1600 variables (some of them are the bits of the wanted
key) -- equation systems of that size are in practice not solvable. This special
equation system is relatively sparse, so only very few of the quadratic terms
(there are about 1,280,000 are possible quadratic terms in total) appear in the
equation system.

The mathematicians Courtois and Pieprzyk \cite{Courtois2002} published a paper
in 2002, which got a great deal of attention amongst the crypto community: The
pair had further developed the XL-method (eXtended Linearization), introduced at
Eurocrypt 2000 by Shamir et al., to create the so called XSL-method (eXtended
Sparse Linearization). The XL-method is a heuristic technique, which in some
cases manages to solve big non-linear equation systems and which was till then
used to analyze an asymmetric algorithm (HFE).  The innovation of Courois and
Pieprzyk was, to apply the XL-method on symmetric codes: the XSL-method can be
applied to very specific equation systems. A 256-bit AES could be attacked in
roughly $2^{230}$ steps. This is still a purely academic attack, but also a
direction pointer for a complete class of block ciphers. The major problem with
this attack is that until now nobody has worked out, under what conditions it is
successful: the authors specify in their paper necessary conditions, but it is
not known, which conditions are sufficient.  There are two very new aspects of
this attack: firstly this attack is not based on statistics but on algebra. So
attacks seem to be possible, where only very small amounts of cipher text are
available. Secondly the security of a product-algorithm does not exponentially
increase with the number of rounds.

>>>>

% $Id$
% ..............................................................................
%                V E R S C H L U E S S E L U N G S V E R F A H R E N
%
% Schreibregelung: Capitel header with all nouns/verbs-words capitalized,
%                  All other headers in the normal lower/upper case manner
%                  like sentences without dot at the end.
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

In their 2001-paper Ferguson, Schroeppel and Whiting \cite{Ferguson2001}
presented a new method of symmetric codes cryptanalysis: They described AES with
a closed formula (in the form of a continued fraction) which was possible
because of the "relatively" clear structure of AES. This formula consists of
around 1000 trillion terms of a sum - so it does not help concrete practical
cryptanalysis. Nevertheless curiosity in the academic community was awakened.
It was already known, that the 128-bit AES could be described as an
over-determined system of about 8000 quadratic equations (over an algebraic
number field) with about 1600 variables (some of them are the bits of the wanted
key) -- equation systems of that size are in practice not solvable. This special
equation system is relatively sparse, so only very few of the quadratic terms
(there are about 1,280,000 are possible quadratic terms in total) appear in the
equation system.

The mathematicians Courtois and Pieprzyk \cite{Courtois2002} published a paper
in 2002, which got a great deal of attention amongst the crypto community: The
pair had further developed the XL-method (eXtended Linearization), introduced at
Eurocrypt 2000 by Shamir et al., to create the so called XSL-method (eXtended
Sparse Linearization). The XL-method is a heuristic technique, which in some
cases manages to solve big non-linear equation systems and which was till then
used to analyze an asymmetric algorithm (HFE).  The innovation of Courtois and
Pieprzyk was, to apply the XL-method on symmetric codes: the XSL-method can be
applied to very specific equation systems. A 256-bit AES could be attacked in
roughly $2^{230}$ steps. This is still a purely academic attack, but also a
direction pointer for a complete class of block ciphers. The major problem with
this attack is that until now nobody has worked out, under what conditions it is
successful: the authors specify in their paper necessary conditions, but it is
not known, which conditions are sufficient.  There are two very new aspects of
this attack: firstly this attack is not based on statistics but on algebra. So
attacks seem to be possible, where only very small amounts of cipher text are
available. Secondly the security of a product-algorithm does not exponentially
increase with the number of rounds.

### 2003-04-08 script/en/chapters/ellipticcurves.tex
be updates

   \item \hypertarget{Lenstra2}{}
         An informative web page about factorisation with elliptic curves. \\
         \href{http://www.loria.fr/~zimmerma/records/ecmnet.html}
	 {\texttt{http://www.loria.fr/\~{}zimmerma/records/ecmnet.html}} \\
         There, one finds literature to the topic factorisation with 
	 elliptic curves as well as links to other web page.

>>>>

   \item \hypertarget{Lenstra2}{}
         An informative web page about factorisation with elliptic curves. \\
         \href{http://www.loria.fr/~zimmerma/records/ecmnet.html}
	 {\texttt{http://www.loria.fr/\~{}zimmerma/records/ecmnet.html}} \\
         It contains literature related to the topic factorisation with 
	 elliptic curves as well as links to other web page.

### 2003-04-08 script/en/chapters/introduction.tex
cvs id
be anmerkungen
be updates

In order to also enable developers/authors to work together independently 
the topics have been split up and for each topic an extra chapter has been 
written which an be read on its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes describing where you
can find the according 
functions within the CrypTool\index{CrypTool} program. Naturally 
there are much more interesting topics in mathematics and cryptography 
which could be discussed in greater depth -- therefore this is only one of many
ways to do it.

@@

I am deeply grateful to all the people helping with their impressive
commitment who have made this global project so successful.

>>>>

In order to also enable developers/authors to work together independently 
the topics have been split up and for each topic an extra chapter has been 
written which can be read on its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes describing where you
can find the according 
functions within the CrypTool\index{CrypTool} program. Naturally 
there are much more interesting topics in mathematics and cryptography 
which could be discussed in greater depth -- therefore this is only one of many
ways to do it.

@@

I am deeply grateful to all the people helping with their impressive
commitment who have made this global project so successful. Especially I would
like to acknowledge the English language proof-reading of this script version
done by Richard Chistenson and L. Montgomery.

### 2003-04-08 script/en/chapters/numbertheory.tex
be updates

% --------------------------------------------------------------------------
\newpage
\subsubsection{Convention}
Unless stated otherwise: 
\begin{itemize}
\item The letter $a, b, c, d, e, k, n, m, p, q$ are used to present whole numbers.
\item The letter $i$ and $j$ represent natural numbers. 
\item The letter $p$ always represents a prime number.
\item Te sets $\mathbb{N} = \{ 1, 2, 3, \cdots \}$ and $\mathbb{Z} =\{ \cdots, -3, -2, -1, 0, 1, 2, 3, \cdots \}$ 
are the {\em natural} and {\em whole} numbers respectively.
\end{itemize}

@@

The numbers that result in the same remainder $r$ when divided by $7$ are combined to form the ``remainder 
class $r$ modulo $7$''. Two numbers $a$ and $b$ belonging to the same remainder class modulo $7$ are said to 
be ``congruent modulo 7''. Or in general:

@@

From algebra theorems it follows that essential parts of the conventional calculation rules are kept when we 
proceed to modular calculations over a basic set $\mathbb{Z}$.  For example, addition remains commutative. 
The same goes for multiplication modulo $m$. The result of a division\footnote{\label{ftn-zth-divmodn}When dividing modulo $m$\index{Division modulo $n$}  we cannot use every number because some numbers have the same 
property as zero. See footnote \ref{ftn-zth-closed}.} is not a fraction but rather a whole number between $0$ and $m-1$.

@@

\begin{itemize}
\item the addition table\footnote{Comment on subtraction modulo 5: \\
$2 - 4 \equiv -2 \equiv 3{\rm ~mod~}5.$\\
It is therefore not true modulo $5$ that $-2 = 2$ (see also \hyperlink{Appendix_C}{Appendix C}). }
${\rm ~(mod~ } 5)$ and
\item the multiplication tables\footnote{Comment on division modulo $6$:\index{Division modulo $n$}\\
Due to the special role of zero as the identity for addition, division by zero is not permitted:\\
for all $a$ it is $a*0=0,$ because $a*0=a*(0+0)=a*0+a*0.$ Obviously $0$ has no inverse regarding the
multiplication, because if there would be one, it must be $ 0= 0*0^{-1} =1.$ Also see footnote \ref{ftn-zth-divmodn}.}
for mod $5$ and mod $6$.
\end{itemize}

@@

\begin{theorem}\label{thm-zth-ordp}\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table 5 for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table 6 demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,21$ are of
    order 22.
  } For a modulus $n$ and $a$ relative prime to
  $n$ the following holds: \\
  $\{ a^i (\mbox{mod }n) | i = 1,\dots,J(n)\}$ equals the multiplicative
  group $Z_n^*$ if and only if $ord_n(a) = J(n)$.
\end{theorem}

@@

\begin{minipage}{\textwidth}
\vspace{3ex}
Links:
\vspace{-10pt}
\begin{itemize}
\item[]   \href{http://www.loria.fr/~zimmerma/records/gnfs158}{\texttt{http://www.loria.fr/~zimmerma/records/gnfs158}} \\
          \href{http://www.crypto-world.com/FactorRecords.html}{\texttt{http://www.crypto-world.com/FactorRecords.html}} \\
          \href{http://www.crypto-world.com/announcements/c158.txt}{\texttt{http://www.crypto-world.com/announcements/c158.txt}}
\end{itemize}
\end{minipage}

As you notice the factorized compound numbers built of 2 prime factors are
much smaller than the especially structured numbers, for which primality 
tests\index{Primality testing} are able to decide whether they are prime 
or not (see chapter 2.4 and 2.5).

@@

\bibitem[Eckert2003]{Eckert2003} \index{Eckert 2003}
    Claudia Eckert, \\
    {\em IT-Sicherheit: Konzepte-Verfahren-Protokolle}, 
    Oldenburg 2001, 2nd edition 2003.

>>>>

% --------------------------------------------------------------------------
\newpage
\subsubsection{Convention}
Unless stated otherwise: 
\begin{itemize}
\item The letter $a, b, c, d, e, k, n, m, p, q$ are used to present whole numbers.
\item The letter $i$~\mbox{and} $j$ represent natural numbers. 
\item The letter $p$ always represents a prime number.
\item Te sets $\mathbb{N} = \{ 1, 2, 3, \cdots \}$ and $\mathbb{Z} =\{ \cdots, -3, -2, -1, 0, 1, 2, 3, \cdots \}$ 
are the {\em natural} and {\em whole} numbers respectively.
\end{itemize}

@@

The numbers that result in the same remainder $r$ when divided by $7$ are combined to form the ``remainder 
class $r$ modulo $7$''. Two numbers $a$ and $b$~\mbox{belonging} to the same remainder class modulo $7$ are said to 
be ``congruent modulo 7''. Or in general:

@@

From algebra theorems it follows that essential parts of the conventional calculation rules are kept when we 
proceed to modular calculations over a basic set $\mathbb{Z}$.  For example, addition remains commutative. 
The same goes for multiplication modulo $m$. The result of a division\footnote{\label{ftn-zth-divmodn}When dividing modulo $m$\index{Division modulo $n$}  we cannot use every number because some numbers have the same 
property as zero. See footnote \ref{ftn-mod6}.} is not a fraction but rather a whole number between $0$ and $m-1$.

@@

\begin{itemize}
\item the addition table\footnote{Comment on subtraction modulo 5: \\
$2 - 4 \equiv -2 \equiv 3{\rm ~mod~}5.$\\
It is therefore not true modulo $5$ that $-2 = 2$ (see also \hyperlink{Appendix_C}{Appendix C}). }
${\rm ~(mod~ } 5)$ and
\item the multiplication tables\footnote{\label{ftn-mod6}Comment on division modulo $6$:\index{Division modulo $n$}\\
Due to the special role of zero as the identity for addition, division by zero is not permitted:\\
for all $a$ it is $a*0=0,$ because $a*0=a*(0+0)=a*0+a*0.$ Obviously $0$ has no inverse regarding the
multiplication, because if there would be one, it must be $ 0= 0*0^{-1} =1.$ Also see footnote \ref{ftn-zth-divmodn}.}
for mod $5$ and mod $6$.
\end{itemize}

@@

\begin{theorem}\label{thm-zth-ordp}\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table 5 for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table 6 demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,19\mbox{ and }21$ are of
    order 22.
  } For a modulus $n$ and $a$ relative prime to
  $n$ the following holds: \\
  $\{ a^i (\mbox{mod }n) | i = 1,\dots,J(n)\}$ equals the multiplicative
  group $Z_n^*$ if and only if $ord_n(a) = J(n)$.
\end{theorem}

@@

\begin{minipage}{\textwidth}
\vspace{3ex}
Links:
\vspace{-10pt}
\begin{itemize}
\item[]   \href{http://www.loria.fr/~zimmerma/records/gnfs158}{\texttt{http://www.loria.fr/~zimmerma/records/gnfs158}} \\
          \href{http://www.crypto-world.com/FactorRecords.html}{\texttt{http://www.crypto-world.com/FactorRecords.html}} \\
          \href{http://www.crypto-world.com/announcements/c158.txt}{\texttt{http://www.crypto-world.com/announcements/c158.txt}}
\end{itemize}
\end{minipage}
\vspace{12pt}

As you notice the factorized compound numbers built of 2 prime factors are
much smaller than the especially structured numbers, for which primality 
tests\index{Primality testing} are able to decide whether they are prime 
or not (see chapters 2.4 and 2.5).

@@

\bibitem[Eckert2003]{Eckert2003} \index{Eckert 2003}
    Claudia Eckert, \\
    {\em IT-Sicherheit: Konzepte-Verfahren-Protokolle}, 
    Oldenbourg 2001, 2nd edition 2003.

### 2003-04-08 script/en/chapters/primes.tex
be updates

\begin{theorem}\label{thm-pz-prod}
Each element $n$ of the natural numbers greater than $1$ can be written as the
product $n = p_1 \cdot p_2 \dots p_m$ of prime numbers. If two such
factorisations $$n =  p_1 \cdot p_2 \cdot \cdots \cdot p_m = p'_1 \cdot p'_2 \cdots
p'_{m'}$$ are given, then they can be reordered such that $\;m = m'\;$ and for
all $i$:  $\;p_i = p'_i$. \\
($n = p_1 \cdot p_2 \cdot \dots \cdot p_m$ are called the prime factors of n)

@@

However, if we know something about the {\em construction} of the number in
question, there are extremely highly developed procedures that are much quicker.
These procedures can determine the primality attribute of a number, but they
cannot determine the prime factors of the numbers, if they are compound.

@@

\vskip +15 pt
\hypertarget{FermatNumbers02}{}
\item {\em Fermat numbers}\footnote{The Fermat prime numbers play a role in circle division.
As proven by Gauss\index{Gauss}, a regular $p$-edge
can only be constructed with the use of a pair of compasses and a ruler, when $p$ is a Fermat prime number.}
\index{Number!Fermat}\index{Fermat!number}  $F(n) = 2^{2^n} + 1$:\\
As mentioned \hyperlink{FermatNumbers01}{above}, Fermat wrote to Mersenne regarding his assumption, that all numbers of this type are primes.
Surprisingly he would have been able obtain a positive result using the negative
prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = 65,537 &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = 4 ,294, 967, 297 = 641 \cdot
6,700,417 &  \mapsto {\rm ~NOT~prime} !
\end{array} 
$$

@@

\pagebreak
  \item $f(n) = n^2 - 79 \cdot n + 1,601$: \\
    This function delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
difference prime values (the same ones as the function from point 11).
$$
\begin{array}{|ll||ll|}
\hline
f(0) = 1,601    & \mapsto {\rm ~prime} &  f(28) = 173    & \mapsto {\rm ~prime}
\\
f(1) = 1,523    & \mapsto {\rm ~prime} &  f(29) = 151    & \mapsto {\rm ~prime}
\\
f(2) = 1,447    & \mapsto {\rm ~prime} &  f(30) = 131 & \mapsto {\rm ~prime} \\
f(3) = 1,373    & \mapsto {\rm ~prime} &  f(31) = 113 & \mapsto {\rm ~prime} \\
f(4) = 1,301    & \mapsto {\rm ~prime} &  f(32) = 97 & \mapsto {\rm ~prime} \\
f(5) = 1,231    & \mapsto {\rm ~prime} &  f(33) = 83 & \mapsto {\rm ~prime} \\
f(6) = 1,163    & \mapsto {\rm ~prime} &  f(34) = 71 & \mapsto {\rm ~prime} \\
f(7) = 1,097    & \mapsto {\rm ~prime} &  f(35) = 61 & \mapsto {\rm ~prime} \\
f(8) = 1,033    & \mapsto {\rm ~prime} &  f(36) = 53 & \mapsto {\rm ~prime} \\
f(9) = 971  & \mapsto {\rm ~prime} &  f(37) = 47 & \mapsto {\rm ~prime} \\
f(10) = 911 & \mapsto {\rm ~prime} &  f(38) = 43 & \mapsto {\rm ~prime} \\
f(11) = 853 & \mapsto {\rm ~prime} &  f(39) = 41 & \mapsto {\rm ~prime} \\
f(12) = 797 & \mapsto {\rm ~prime} &  f(40) = 41 & \mapsto {\rm ~prime} \\               
f(13) = 743 & \mapsto {\rm ~prime} &  f(41) = 43 & \mapsto {\rm ~prime} \\               
f(14) = 691 & \mapsto {\rm ~prime} &  f(42) = 47 & \mapsto {\rm ~prime} \\               
f(15) = 641 & \mapsto {\rm ~prime} &  f(43) = 53 & \mapsto {\rm ~prime} \\               
f(16) = 593 & \mapsto {\rm ~prime} &  \cdots  &  \\                                      
f(17) = 547 & \mapsto {\rm ~prime} &  f(77) = 1,447  & \mapsto {\rm ~prime} \\           
f(18) = 503 & \mapsto {\rm ~prime} &  f(78) = 1,523  & \mapsto {\rm ~prime} \\           
f(19) = 461 & \mapsto {\rm ~prime} &  f(79) = 1,601  & \mapsto {\rm ~prime} \\           
f(20) = 421 & \mapsto {\rm ~prime} &  f(80) = 11 \cdot 151 & \mapsto {\rm ~NOT~prime!} \\
f(21) = 383 & \mapsto {\rm ~prime} &  f(81) = 41 \cdot 43 & \mapsto {\rm ~NOT~prime!} \\ 
f(22) = 347 & \mapsto {\rm ~prime} &  f(82) = 1,847  & \mapsto {\rm ~prime} \\           
f(21) = 383 & \mapsto {\rm ~prime} &  f(83) = 1,933  & \mapsto {\rm ~prime} \\           
f(22) = 347 & \mapsto {\rm ~prime} &  f(84) = 43 \cdot 47 &  \mapsto {\rm ~NOT~prime!} \\
f(23) = 313 & \mapsto {\rm ~prime} & & \\
f(24) = 281 & \mapsto {\rm ~prime} & & \\
f(25) = 251 & \mapsto {\rm ~prime} & & \\
f(26) = 223 & \mapsto {\rm ~prime} & & \\
f(27) = 197 & \mapsto {\rm ~prime} & & \\
\hline
\end{array} 
$$

@@

\pagebreak
  \item  \index{Catalan number} Catalan, after whom the so-called
{\em Catalan numbers} $A(n) = (1 / (n+1) ) * (2n)! / (n!)^2$ are named,
conjectured that $C_4$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

The following proves that there are more even numbers than square ones:
\begin{itemize}
  \item the size of the $n$th element: \\
    The $n$th element of the even numbers is $2n$; the $n$th element of the
square numbers is $n^2$. Because for all $n>2$: $2n < n^2$, the $n$th even
number occurs much earlier than the $n$th square number.
    Thus the even numbers are distributed more densely and we can say that there
are more even numbers than square ones.
  \item the number of values that are less than or equal to a certain {\em
maximum value} $x$ in ${\mathbb R}$ is: \\
    There are $[x/2]$ such even numbers and $[\sqrt{x}]$ square numbers. Because
for large $x$ the value $x/2$ is much greater than the square root of $2$, we
can again say that there are more even numbers.
\end{itemize}
\vskip +3 pt
\begin{theorem}\label{thm-pz-density}
For large n: The value of the $n$th prime $P(n)$ is asymptotic to $n \cdot
ln(n)$, i.e. the limit of the relation  $P(n)/(n\cdot \ln n)$ is equal to $1$ if
$n$ tends to infinity.
\end{theorem}

@@

\bibitem[Tietze1973]{Tietze1973} \index{Tietze 1973}     
    H. Tietze, \\
    {\em Gel\"oste und ungel\"oste mathematische Probleme}, 
    Verlag C.H. Beck 1959, 6th edition 1973.

>>>>

\begin{theorem}\label{thm-pz-prod}
Each element $n$ of the natural numbers greater than $1$ can be written as the
product $n = p_1 \cdot p_2 \dots p_m$ of prime numbers. If two such
factorisations $$n =  p_1 \cdot p_2 \cdot \cdots \cdot p_m = p'_1 \cdot p'_2 \cdots
p'_{m'}$$ are given, then they can be reordered such that $\;m = m'\;$ and for
all $i$:  $\;p_i = p'_i$. \\
($p_1, p_2, \dots, p_m$ are called the prime factors of n)

@@

However, if we know something about the {\em construction} of the number in
question, there are extremely highly developed procedures that are much quicker.
These procedures can determine the primality attribute of a number, but they
cannot determine the prime factors of a number, if it is compound.

@@

\vskip +15 pt
\hypertarget{FermatNumbers02}{}
\item {\em Fermat numbers}\footnote{The Fermat prime numbers play a role in circle division.
As proven by Gauss\index{Gauss}, a regular $p$-edge
can only be constructed with the use of a pair of compasses and a ruler, when $p$ is a Fermat prime number.}
\index{Number!Fermat}\index{Fermat!number}  $F(n) = 2^{2^n} + 1$:\\
As mentioned \hyperlink{FermatNumbers01}{above}, Fermat wrote to Mersenne regarding his assumption, that all numbers of this type are primes.
Surprisingly he would have been able obtain a positive result using the negative
prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
F(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !
\end{array} 
$$

@@

  \item $f(n) = n^2 - 79 \cdot n + 1,601$: \\
    This function delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
difference prime values (the same ones as the function from point 11).
$$
\begin{array}{|ll||ll|}
\hline
f(0) = 1,601    & \mapsto {\rm ~prime} &  f(28) = 173    & \mapsto {\rm ~prime}
\\
f(1) = 1,523    & \mapsto {\rm ~prime} &  f(29) = 151    & \mapsto {\rm ~prime}
\\
f(2) = 1,447    & \mapsto {\rm ~prime} &  f(30) = 131 & \mapsto {\rm ~prime} \\
f(3) = 1,373    & \mapsto {\rm ~prime} &  f(31) = 113 & \mapsto {\rm ~prime} \\
f(4) = 1,301    & \mapsto {\rm ~prime} &  f(32) = 97 & \mapsto {\rm ~prime} \\
f(5) = 1,231    & \mapsto {\rm ~prime} &  f(33) = 83 & \mapsto {\rm ~prime} \\
f(6) = 1,163    & \mapsto {\rm ~prime} &  f(34) = 71 & \mapsto {\rm ~prime} \\
f(7) = 1,097    & \mapsto {\rm ~prime} &  f(35) = 61 & \mapsto {\rm ~prime} \\
f(8) = 1,033    & \mapsto {\rm ~prime} &  f(36) = 53 & \mapsto {\rm ~prime} \\
f(9) = 971  & \mapsto {\rm ~prime} &  f(37) = 47 & \mapsto {\rm ~prime} \\
f(10) = 911 & \mapsto {\rm ~prime} &  f(38) = 43 & \mapsto {\rm ~prime} \\
f(11) = 853 & \mapsto {\rm ~prime} &  f(39) = 41 & \mapsto {\rm ~prime} \\
f(12) = 797 & \mapsto {\rm ~prime} &  f(40) = 41 & \mapsto {\rm ~prime} \\               
f(13) = 743 & \mapsto {\rm ~prime} &  f(41) = 43 & \mapsto {\rm ~prime} \\               
f(14) = 691 & \mapsto {\rm ~prime} &  f(42) = 47 & \mapsto {\rm ~prime} \\               
f(15) = 641 & \mapsto {\rm ~prime} &  f(43) = 53 & \mapsto {\rm ~prime} \\               
f(16) = 593 & \mapsto {\rm ~prime} &  \cdots  &  \\                                      
f(17) = 547 & \mapsto {\rm ~prime} &  f(77) = 1,447  & \mapsto {\rm ~prime} \\           
f(18) = 503 & \mapsto {\rm ~prime} &  f(78) = 1,523  & \mapsto {\rm ~prime} \\           
f(19) = 461 & \mapsto {\rm ~prime} &  f(79) = 1,601  & \mapsto {\rm ~prime} \\           
f(20) = 421 & \mapsto {\rm ~prime} &  f(80) = 11 \cdot 151 & \mapsto {\rm ~NOT~prime!} \\
f(21) = 383 & \mapsto {\rm ~prime} &  f(81) = 41 \cdot 43 & \mapsto {\rm ~NOT~prime!} \\ 
f(22) = 347 & \mapsto {\rm ~prime} &  f(82) = 1,847  & \mapsto {\rm ~prime} \\           
f(21) = 383 & \mapsto {\rm ~prime} &  f(83) = 1,933  & \mapsto {\rm ~prime} \\           
f(22) = 347 & \mapsto {\rm ~prime} &  f(84) = 43 \cdot 47 &  \mapsto {\rm ~NOT~prime!} \\
f(23) = 313 & \mapsto {\rm ~prime} & & \\
f(24) = 281 & \mapsto {\rm ~prime} & & \\
f(25) = 251 & \mapsto {\rm ~prime} & & \\
f(26) = 223 & \mapsto {\rm ~prime} & & \\
f(27) = 197 & \mapsto {\rm ~prime} & & \\
\hline
\end{array} 
$$

@@

  \item  \index{Catalan number} Catalan, after whom the so-called
{\em Catalan numbers} $A(n) = (1 / (n+1) ) * (2n)! / (n!)^2$ are named,
conjectured that $C_4$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

The following proves that there are more even numbers than square ones:
\begin{itemize}
  \item the size of the $n$th element: \\
    The $n$th element of the even numbers is $2n$; the $n$th element of the
square numbers is $n^2$. Because for all $n>2$: $2n < n^2$, the $n$th even
number occurs much earlier than the $n$th square number.
    Thus the even numbers are distributed more densely and we can say that there
are more even numbers than square ones.
  \item the number of values that are less than or equal to a certain {\em
maximum value} $x$ in ${\mathbb R}$ is: \\
    There are $[x/2]$ such even numbers and $[\sqrt{x}]$ square numbers. Because
for large $x$ the value $x/2$ is much greater than the square root of $2$, we
can again say that there are more even numbers.
\end{itemize}
\vskip +3 pt
\begin{theorem}\label{thm-pz-density}
For large n: The value of the $n$-th prime $P(n)$ is asymptotic to $n \cdot
ln(n)$, i.e. the limit of the relation  $P(n)/(n\cdot \ln n)$ is equal to $1$ if
$n$ tends to infinity.
\end{theorem}

@@

\bibitem[Tietze1973]{Tietze1973} \index{Tietze 1973}     
    H. Tietze, \\
    {\em Gel\"oste und ungel\"oste mathematische Probleme}, \\
    Verlag C. H. Beck 1959, 6th edition 1973.
